---
title: "All Parts"
author: "Gabriella Weis"
date: "2026-03-31"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Libraries and Data

```{r,message=FALSE,warning=FALSE}
library(tidyverse)
library(GGally)
library(caret)
library(rpart)
library(rpart.plot)
library(class)
library(ipred)
library(randomForest)
library(vip)
library(xgboost)
library(yardstick)
library(colorspace)
library(stats)
library(factoextra)
library(tidyr)
```

```{r}
#SUBSET, 150 OBS (50 EACH OBJ)
galaxy_sub <- read_csv("galaxy.csv")
qso_sub <- read_csv("qso.csv")
star_sub <- read_csv("star.csv")

#LARGE SUBSET, 3000 OBS (1000 EACH OBJ)
galaxy <- read_csv("ALLGALAXY.csv")
qso <- read_csv("ALLQSO.csv")
star <- read_csv("ALLSTAR.csv")
```

## Wrangling

```{r}
ds_sub=star_sub %>%
  rbind(galaxy_sub) %>%
  rbind(qso_sub) %>%
  select(-specobjid,-objid,-redshift,-ra,-dec) %>%
  mutate(class=as.factor(class)) %>%
  mutate(`u-g`=u-g,`g-r`=g-r,`r-i`=r-i,`i-z`=i-z)
#head(ds_sub)
```

```{r}
ds=star %>%
  rbind(galaxy) %>%
  rbind(qso) %>%
  select(-specobjid,-objid,-redshift,-ra,-dec) %>%
  mutate(class=as.factor(class)) %>%
  mutate(`u-g`=u-g,`g-r`=g-r,`r-i`=r-i,`i-z`=i-z)
head(ds)
```

## GGPairs

```{r}
ggpairs(ds_sub, aes(color=class), columns=c("u-g","g-r","r-i","i-z"))
ggpairs(ds, aes(color=class), columns=c("u-g","g-r","r-i","i-z"))
```

## Variables (all unitless)

| Variable Name | Type | Definition |
|------------------------|------------------------|------------------------|
| class | categorical (3 levels) | category of object (star, quasar, galaxy) |
| redshift | numeric | how fast an object is moving away from us |
| u-g | numeric | color scale calculated from flux magnitudes u and g |
| g-r | numeric | color scale calculated from flux magnitudes g and r |
| r-i | numeric | color scale calculated from flux magnitudes r and i |
| i-z | numeric | color scale calculated from flux magnitudes i and z |

**Important Note**: During initial KNN and Classification Tree model creation, we noticed that redshift was too prevalent of a predictor, and often unatainable for use in real life. Therefore, we chose to exclude it from the models for the time being.

## KNN and Classification Tree

### Testing and Training Splits

```{r}
set.seed(1)

caretSamp <- createDataPartition(ds$class , 
                                        p = 0.7, 
                                        list = FALSE)

## SPLIT TESTING AND TRAINING
trainCaret  <- ds[caretSamp, ]
testCaret <- ds[-caretSamp, ]
```

### KNN

```{r}
set.seed(1)

model <- train(
  class ~., 
  data = trainCaret, 
  method = "knn",
  trControl = trainControl("cv", number = 10),
  preProcess = c("range"),
  tuneLength = 20
)
# Plot model accuracy vs different values of k
plot(model)

# Print the best tuning parameter k that
# maximizes model accuracy
model$bestTune

predicted.classes <- model %>% predict(testCaret)
head(predicted.classes)

# Compute model accuracy rate
cmCaret<-table(predicted.classes, testCaret$class)
#cmCaret

### THE CARET WAY!
confusionMatrix(cmCaret)
```

**Explanation**: The best value of k is 5, as determined by cross validation. The confusion matrix and accuracy rating (almost 90%) shows that KNN produces a relatively good model. We noticed that it had some difficulty differentiating between galaxies and stars, which proved a theme across models.

### Basic Tree

```{r}
set.seed(1)

classTree<- rpart(class ~ .,
                  data = trainCaret,
                  method = "class"
                  )

### PLOT TREE
rpart.plot(classTree)

### Plot CP
plotcp(classTree)

printcp(classTree)

## WHICH CP
minCP<-classTree$cptable[which.min(classTree$cptable[,"xerror"]),"CP"]
minCP
```

### Pruned Tree

```{r}
set.seed(1)

prune_classTree <- rpart(class ~ .,
                         data = trainCaret,
                         method = "class",
                         control = rpart.control(cp=minCP)
                         )

rpart.plot(prune_classTree)
```

### Unpruned vs. Pruned Predictions (Same)

```{r}
## DEFAULT TREE 
### PREDICT
predTree1<-predict(classTree, testCaret, type="class")

### CONFUSION MATRIX
cmTree1<-table(testCaret$class, predTree1)
cmTree1

#### CORRECT RATE
mean(testCaret$class==predTree1)

## PRUNED TREE 
### PREDICT
predTree2<-predict(prune_classTree, testCaret, type="class")

### CONFUSION MATRIX
cmTree2<-table(testCaret$class, predTree2)
cmTree2

#### CORRECT RATE
mean(testCaret$class==predTree2)
```

**Explanation**: The classification tree was the same pruned as it was unpruned, and performed similarly to the KNN model, as can be seen by the confusion matrix and the accuracy rating (almost 90%). The star/galaxy classification uncertainty remains a problem in this model.

# Part 2

## Bagging

```{r}
set.seed(1)

pimaBag <- bagging(class ~ .,
                   data = trainCaret,
                   nbagg = 150,   
                   coob = TRUE,
                   control = rpart.control(minsplit = 2, cp = 0))

## PREDICT
predBag<-predict(pimaBag, testCaret, type="class")

## CONFUSION MATRIX
cmBag<-table(testCaret$class, predBag)
cmBag

## CORRECT RATE
mean(testCaret$class==predBag)
```

## Random Forest

```{r}
set.seed(1)

caretRF <- train(class ~., 
               data = trainCaret, 
               method = "rf",
               trControl = trainControl("cv", number = 10),
               importance = TRUE
)
# Best tuning parameter
caretRF$bestTune

# Final model
caretRF$finalModel

## PREDICT
predCaretRF <- caretRF %>% predict(testCaret)

## TABLE
table(predCaretRF, testCaret$class)

## MEAN
mean(predCaretRF == testCaret$class)
```

```{r}
# Plot MeanDecreaseAccuracy
varImpPlot(caretRF$finalModel, type = 1)
# Plot MeanDecreaseGini
varImpPlot(caretRF$finalModel, type = 2)

vip(caretRF$finalModel, type = 1) 
vip(caretRF$finalModel, type = 2)
```

## Boosting

```{r, warning=FALSE}
library(tidymodels)

# Define the model
xgb_model <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("classification")

# Define the workflow
xgb_workflow <- workflow() %>%
  add_model(xgb_model) %>%
  add_formula(class ~.)

# Define 10-fold cross-validation
cv_folds <- vfold_cv(trainCaret, v = 10)

# Fit with resampling
xgb_results <- fit_resamples(
  xgb_workflow,
  resamples = cv_folds,
  control = control_resamples(save_pred = TRUE)
)

# View CV metrics
collect_metrics(xgb_results)

# Fit final model on full training data
final_fit <- fit(xgb_workflow, data = trainCaret)

# Make predictions on the test data
predBoost <- final_fit %>% predict(testCaret)

## TABLE
table(predBoost$.pred_class, testCaret$class)

mean(predBoost$.pred_class == testCaret$class)
```

# Part 3

## Collapse Categorical Variable

```{r}
col_ds=ds %>%
  filter(class!="QSO") %>%
  mutate(class=ifelse(class=="STAR",0,1))

head(col_ds)
```

## Simple Logistic Regression

```{r}
set.seed(1)

caretSamp2 <- createDataPartition(col_ds$class, 
                                        p = 0.7, 
                                        list = FALSE)

## SPLIT TESTING AND TRAINING
trainCaret2  <- col_ds[caretSamp2, ]
testCaret2 <- col_ds[-caretSamp2, ]

modsim <- glm(class~`r-i`, trainCaret2, family="binomial")
summary(modsim)
```

```{r}
ggplot(trainCaret2,aes(x=`r-i`,y=class)) +
  geom_point() +
  geom_line(aes(x=`r-i`,y=modsim$fitted),color="blue")
  
```

### Prediction

```{r}
#prediction
pred1<-predict(modsim,testCaret2,type="response")
head(pred1)
```

```{r}
#confusion matrix
conf_mat1<-data.frame(testOutcome=testCaret2$class, predOut=pred1>.5)
table(conf_mat1$predOut, conf_mat1$testOutcome)

## CORRECT RATE
mean(conf_mat1$predOut==conf_mat1$testOutcome)
```

## Logistic Regression

```{r}
mod1 <- glm(class~`r-i`+`u-g`+`g-r`+`i-z`, trainCaret2, family="binomial")
summary(mod1)
```

### Backward Selection & Prediction

```{r}
stats::step(mod1)
```

```{r}
modbest<-glm(class~`r-i`+`u-g`+`i-z`, trainCaret2, family="binomial")
summary(modbest)
```

```{r}
#prediction
pred2<-predict(modbest,testCaret2,type="response")
head(pred2)
```

```{r}
#confusion matrix
conf_mat2<-data.frame(testOutcome=testCaret2$class, predOut=pred2>.5)
table(conf_mat2$predOut, conf_mat2$testOutcome)

## CORRECT RATE
mean(conf_mat2$predOut==conf_mat2$testOutcome)
```

### Best Model With Fixed Intercept

```{r,warning=FALSE}
library(MuMIn)

full_model <- glm(class ~ ., 
                  data = ds, 
                  family = binomial, 
                  na.action = "na.fail")

all_models <- dredge(full_model, rank = "AIC")

#print(all_models)

best_model <- get.models(all_models, 1)[[1]]
summary(best_model)
```

# Part 4: Extra Model (PCA)

```{r}
pca_ds <- ds[,c("u-g","g-r","r-i","i-z")]
pca <- prcomp(pca_ds,scale=TRUE)
summary(pca)
pca$rotation
```

```{r}
fviz_eig(pca, addlabels = TRUE)
```

```{r}
fviz_pca_var(pca,col.var="steelblue")
fviz_pca_var(pca,col.var="steelblue",axes=c(1,3))
fviz_pca_var(pca,col.var="steelblue",axes=c(2,3))
```

```{r}
groups  <-  ds$class
fviz_pca_ind(pca,
             col.ind = groups, # color by groups
             palette = c("blue","red","green"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups",
             repel = FALSE,
             label="none",
             axes=c(1,2)
)
```
```{r}
#above plot, zoomed in

groups  <-  ds$class
fviz_pca_ind(pca,
             col.ind = groups, # color by groups
             palette = c("blue","red","green"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups",
             repel = FALSE,
             label="none",
             axes=c(1,2),
             xlim = c(-4, 4),
             ylim = c(-3, 3)
)
```

# Part 5: Extra Visualizations

## KNN

```{r}
con_knn=testCaret
con_knn$pred=predicted.classes
```

```{r}
cm <- conf_mat(con_knn, class, pred)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: KNN")
```



## Trees (Pruned & Unpruned)

```{r}
#new predictions column
con_tree=testCaret
con_tree$pred=predTree1
```

```{r}
#heatmap
cm <- conf_mat(con_tree, class, pred)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: Unpruned Tree")
```

```{r}
con_tree2=testCaret
con_tree2$pred=predTree2
```

```{r}
cm <- conf_mat(con_tree2, class, pred)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: Pruned Tree")
```

## Bagging

```{r}
con_bag=testCaret
con_bag$pred=predBag
```

```{r}
cm <- conf_mat(con_bag, class, pred)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: Bagging")
```

## Random Forest

```{r}
con_rf=testCaret
con_rf$pred=predCaretRF
```

```{r}
cm <- conf_mat(con_rf, class, pred)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: Random Forest")
```

## Boosting

```{r}
con_boost=testCaret
con_boost$pred=predBoost$.pred_class
```

```{r}
cm <- conf_mat(con_boost, class, pred)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: Boosting")
```

## Logistic Regression (Best Multiple)

```{r}
con_log=testCaret2
con_log$pred=conf_mat2$predOut
con_log=con_log %>%
  mutate(class=as.factor(case_when(
    class==0 ~ "STAR",
    class==1 ~ "GALAXY")),
    pred=as.factor(pred))
```

```{r}
con_log=con_log %>%
  mutate(pred_var=factor(case_when(
    pred=="FALSE" ~ "STAR",
    pred=="TRUE" ~ "GALAXY"
  )))
```

```{r}
cm <- conf_mat(con_log, class, pred_var)
autoplot(cm, type="heatmap") +
  scale_fill_continuous_sequential(palette="Peach") +
  labs(title="Confusion Matrix: Best Multiple Logistic Regression")
```
