---
title: "Part 1"
author: "Gabriella Weis"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Libraries and Data

```{r,message=FALSE,warning=FALSE}
library(tidyverse)
library(GGally)
library(caret)
library(rpart)
library(rpart.plot)
library(class)
library(ipred)
library(randomForest)
library(vip)
library(xgboost)
```

```{r}
galaxy <- read_csv("galaxy.csv")
qso <- read_csv("qso.csv")
star <- read_csv("star.csv")
```

## Wrangling

```{r}
ds=star %>%
  rbind(galaxy) %>%
  rbind(qso) %>%
  select(-specobjid,-objid,-redshift) %>%
  mutate(class=as.factor(class)) %>%
  mutate(`u-g`=u-g,`g-r`=g-r,`r-i`=r-i,`i-z`=i-z)
head(ds)
```

## GGPairs

```{r}
ggpairs(ds, aes(color=class), columns=c("u-g","g-r","r-i","i-z"))
```

## Variables (all unitless)

| Variable Name | Type | Definition |
|------------------------|------------------------|------------------------|
| class | categorical (3 levels) | category of object (star, quasar, galaxy) |
| redshift | numeric | how fast an object is moving away from us |
| u-g | numeric | color scale calculated from flux magnitudes u and g |
| g-r | numeric | color scale calculated from flux magnitudes g and r |
| r-i | numeric | color scale calculated from flux magnitudes r and i |
| i-z | numeric | color scale calculated from flux magnitudes i and z |

**Important Note**: During initial KNN and Classification Tree model creation, we noticed that redshift was too prevalent of a predictor, and often unatainable for use in real life. Therefore, we chose to exclude it from the models for the time being.

## KNN and Classification Tree

### Testing and Training Splits

```{r}
set.seed(1)

caretSamp <- createDataPartition(ds$class , 
                                        p = 0.7, 
                                        list = FALSE)

## SPLIT TESTING AND TRAINING
trainCaret  <- ds[caretSamp, ]
testCaret <- ds[-caretSamp, ]
```

### KNN

```{r}
set.seed(1)

model <- train(
  class ~., 
  data = trainCaret, 
  method = "knn",
  trControl = trainControl("cv", number = 10),
  preProcess = c("range"),
  tuneLength = 20
)
# Plot model accuracy vs different values of k
plot(model)

# Print the best tuning parameter k that
# maximizes model accuracy
model$bestTune

predicted.classes <- model %>% predict(testCaret)
head(predicted.classes)

# Compute model accuracy rate
cmCaret<-table(predicted.classes, testCaret$class)
#cmCaret

### THE CARET WAY!
confusionMatrix(cmCaret)
```

**Explanation**: The best value of k is 5, as determined by cross validation. The confusion matrix and accuracy rating (over 75%) shows that KNN produces a relatively good model. We noticed that it had some difficulty differentiating between galaxies and stars, which proved a theme across models.

### Basic Tree

```{r}
set.seed(1)

classTree<- rpart(class ~ .,
                  data = trainCaret,
                  method = "class"
                  )

### PLOT TREE
rpart.plot(classTree)

### Plot CP
plotcp(classTree)

printcp(classTree)

## WHICH CP
minCP<-classTree$cptable[which.min(classTree$cptable[,"xerror"]),"CP"]
minCP
```

### Pruned Tree

```{r}
set.seed(1)

prune_classTree <- prune(classTree, cp = minCP)
rpart.plot(prune_classTree)
```

### Unpruned vs. Pruned Predictions (Same)

```{r}
## DEFAULT TREE 
### PREDICT
predTree1<-predict(classTree, testCaret, type="class")

### CONFUSION MATRIX
cmTree1<-table(testCaret$class, predTree1)
cmTree1

#### CORRECT RATE
mean(testCaret$class==predTree1)

## PRUNED TREE 
### PREDICT
predTree2<-predict(prune_classTree , testCaret, type="class")

### CONFUSION MATRIX
cmTree2<-table(testCaret$class, predTree2)
cmTree2

#### CORRECT RATE
mean(testCaret$class==predTree2)
```

**Explanation**: The classification tree was the same pruned as it was unpruned, and performed much better than the KNN model, as can be seen by the confusion matrix and the accuracy rating (over 90%). The star/galaxy classification uncertainty remains a problem, though significantly less so in this model.

# Part 2

## Bagging

```{r}
set.seed(1)

pimaBag <- bagging(class ~ .,
                   data = trainCaret,
                   nbagg = 150,   
                   coob = TRUE,
                   control = rpart.control(minsplit = 2, cp = 0))

## PREDICT
predBag<-predict(pimaBag, testCaret, type="class")

## CONFUSION MATRIX
cmBag<-table(testCaret$class, predBag)
cmBag

## CORRECT RATE
mean(testCaret$class==predBag)
```

## Random Forest

```{r}
set.seed(1)

caretRF <- train(class ~., 
               data = trainCaret, 
               method = "rf",
               trControl = trainControl("cv", number = 10),
               importance = TRUE
)
# Best tuning parameter
caretRF$bestTune

# Final model
caretRF$finalModel

## PREDICT
predCaretRF <- caretRF %>% predict(testCaret)

## TABLE
table(predCaretRF, testCaret$class)

## MEAN
mean(predCaretRF == testCaret$class)
```

```{r}
# Plot MeanDecreaseAccuracy
varImpPlot(caretRF$finalModel, type = 1)
# Plot MeanDecreaseGini
varImpPlot(caretRF$finalModel, type = 2)

vip(caretRF$finalModel, type = 1) 
vip(caretRF$finalModel, type = 2)
```

## Boosting

```{r, warning=FALSE}
library(tidymodels)

# Define the model
xgb_model <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("classification")

# Define the workflow
xgb_workflow <- workflow() %>%
  add_model(xgb_model) %>%
  add_formula(class ~.)

# Define 10-fold cross-validation
cv_folds <- vfold_cv(trainCaret, v = 10)

# Fit with resampling
xgb_results <- fit_resamples(
  xgb_workflow,
  resamples = cv_folds,
  control = control_resamples(save_pred = TRUE)
)

# View CV metrics
collect_metrics(xgb_results)

# Fit final model on full training data
final_fit <- fit(xgb_workflow, data = trainCaret)

# Make predictions on the test data
predBoost <- final_fit %>% predict(testCaret)

## TABLE
table(predBoost$.pred_class, testCaret$class)

mean(predBoost$.pred_class == testCaret$class)
```

# Part 3

## Collapse Categorical Variable

```{r}
col_ds=ds %>%
  filter(class!="QSO") %>%
  mutate(class=ifelse(class=="STAR",0,1))

head(col_ds)
```

## Simple Logistic Regression

```{r}
set.seed(1)

caretSamp <- createDataPartition(col_ds$class, 
                                        p = 0.7, 
                                        list = FALSE)

## SPLIT TESTING AND TRAINING
trainCaret  <- col_ds[caretSamp, ]
testCaret <- col_ds[-caretSamp, ]

modsim <- glm(class~`r-i`, trainCaret, family="binomial")
summary(modsim)
```

```{r}
ggplot(trainCaret,aes(x=`r-i`,y=class)) +
  geom_point() +
  geom_line(aes(x=`r-i`,y=modsim$fitted),color="blue")
  
```

## Logistic Regression

```{r}
mod1 <- glm(class~`r-i`+`u-g`, trainCaret, family="binomial")
summary(mod1)
```
