---
title: "Testing the dataset"
author: "Gabriella Weis"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

## Librarys and Data

```{r}
library(tidyverse)
library(GGally)
library(caret)
library(rpart)
library(rpart.plot)
library(class)
library(ipred)
```

```{r}
galaxy <- read_csv("galaxy.csv")
qso <- read_csv("qso.csv")
star <- read_csv("star.csv")
```

## Wrangling

```{r}
ds=star %>%
  rbind(galaxy) %>%
  rbind(qso) %>%
  select(-specobjid,-objid,-redshift) %>%
  mutate(class=as.factor(class)) %>%
  mutate(`u-g`=u-g,`g-r`=g-r,`r-i`=r-i,`i-z`=i-z)

#ds=cbind(ds[,-8],ds[,8])
head(ds)
```

## GGPairs

```{r}
ggpairs(ds, aes(color=class), columns=c("u-g","g-r","r-i","i-z"))
```

## Testing and Training Splits

```{r}
set.seed(1)

caretSamp <- createDataPartition(ds$class , 
                                        p = 0.7, 
                                        list = FALSE)

## SPLIT TESTING AND TRAINING
trainCaret  <- ds[caretSamp, ]
testCaret <- ds[-caretSamp, ]
```

### KNN

```{r}
set.seed(1)

model <- train(
  class ~., 
  data = trainCaret, 
  method = "knn",
  trControl = trainControl("cv", number = 10),
  preProcess = c("range"),
  tuneLength = 20
)
# Plot model accuracy vs different values of k
plot(model)

# Print the best tuning parameter k that
# maximizes model accuracy
model$bestTune

predicted.classes <- model %>% predict(testCaret)
head(predicted.classes)

# Compute model accuracy rate
cmCaret<-table(predicted.classes, testCaret$class)
cmCaret

### THE CARET WAY!
confusionMatrix(cmCaret)
```

### Tree

```{r}
set.seed(1)

classTree<- rpart(class ~ .,
                  data = trainCaret,
                  method = "class"
                  )

### PLOT TREE
rpart.plot(classTree)

### Plot CP
plotcp(classTree)

printcp(classTree)

## WHICH CP
minCP<-classTree$cptable[which.min(classTree$cptable[,"xerror"]),"CP"]
minCP
```

### Pruned Tree

```{r}
set.seed(1)

prune_classTree <- prune(classTree, cp = minCP)
rpart.plot(prune_classTree)
```

### Unpruned vs. Pruned Predictions (Same)

```{r}
## DEFAULT TREE 
### PREDICT
predTree1<-predict(classTree, testCaret, type="class")

### CONFUSION MATRIX
cmTree1<-table(testCaret$class, predTree1)
cmTree1

#### CORRECT RATE
mean(testCaret$class==predTree1)

## PRUNED TREE 
### PREDICT
predTree2<-predict(prune_classTree , testCaret, type="class")

### CONFUSION MATRIX
cmTree2<-table(testCaret$class, predTree2)
cmTree2

#### CORRECT RATE
mean(testCaret$class==predTree2)
```

## Variables (all unitless)

| Variable Name | Type | Definition |
|----|----|----|
| class | categorical (3 levels) | category of object (star, quasar, galaxy) |
| redshift | numeric | how fast an object is moving away from us |
| u-g | numeric |  |
| g-r | numeric |  |
| r-i | numeric |  |
| i-z | numeric |  |

# Part 2

## Bagging

```{r}
set.seed(1)

caretBag <- train(class ~., 
               data = trainCaret, 
               method = "treebag",
               trControl = trainControl("cv", number = 5),
               importance = TRUE
)

predCaretBag <- caretBag %>% predict(testCaret)

# CONFUSION MATRIX
table(predCaretBag, testCaret$class)

# CORRECT RATE
mean(predCaretBag == testCaret$class)

```

```{r}
set.seed(1)

pimaBag <- bagging(class ~ .,
                   data = trainCaret,
                   nbagg = 150,   
                   coob = TRUE,
                   control = rpart.control(minsplit = 2, cp = 0))

## PREDICT
predBag<-predict(pimaBag, testCaret, type="class")

## CONFUSION MATRIX
cmBag<-table(testCaret$class, predBag)
cmBag

## CORRECT RATE
mean(testCaret$class==predBag)
```

## Random Forest

```{r}

```
